Escape Analysis and Specialization in a Tracing JIT

### 2010-09-28 talk/pepm2011/paper.tex
fix some XXXs

The basic approach of virtual objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:XXX} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.

@@

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.\footnote{In theory, the
    approach of Section~\ref{sec:XXX} works also for objects that live for
    exactly $n>1$ iterations and then don't escape, but we expect this to be a
    very rare case, so we do not handle it.}
\end{itemize}

@@

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object\footnote{XXX what I have in mind when I talk
of ``virtual object'' is the run-time behavior -- i.e. a real object that
would exist at run-time, except that it has be virtual-ized.  Here you seem
to mean rather ``virtual object description'' or something.}
is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the optimization to improve the trace. The shapes describe
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.

@@

XXX ``the trace above'' is dangerous; should mention all figures by numbers

This is what happens at the end of the trace above, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point virtual objects. Before the
jump is emitted, they are \emph{forced}. This means that the optimizers produces code
that allocates a new object of the right type and sets its fields to the field
values that the virtual object has. This means that instead of the jump, the
following operations are emitted:

@@

The reason why we cannot optimize the remaining allocations away is because
their lifetime crosses the jump. To improve the situation, a little trick is
needed. The trace above represents a loop, i.e. the jump at the end jumps to
the beginning. Where in the loop the jump occurs is arbitrary, since the loop
can only be left via failing guards anyway. Therefore it does not change the
semantics of the loop to put the jump at another point into the trace and we
can move the \texttt{jump} operation just above the allocation of the objects that
appear in the current \texttt{jump}. This needs some care, because the arguments to
\texttt{jump} are all currently live variables, thus they need to be adapted.

If we do that for our example trace above, the trace looks like this:
\begin{figure}
\includegraphics{figures/step2.pdf}
\end{figure}

@@

>>>>

The basic approach of virtual objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:support} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.

@@

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.\footnote{In theory, the
    approach of Section~\ref{sec:crossloop} works also for objects that live for
    exactly $n>1$ iterations and then don't escape, but we expect this to be a
    very rare case, so we do not handle it.}
\end{itemize}

@@

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object\arigo{XXX what I have in mind when I talk
of ``virtual object'' is the run-time behavior -- i.e. a real object that
would exist at run-time, except that it has be virtual-ized.  Here you seem
to mean rather ``virtual object description'' or something.}
is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the optimization to improve the trace. The shapes describe
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.

@@

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point virtual objects. Before the
jump is emitted, they are \emph{forced}. This means that the optimizers produces code
that allocates a new object of the right type and sets its fields to the field
values that the virtual object has. This means that instead of the jump, the
following operations are emitted:

@@

The reason why we cannot optimize the remaining allocations away is because
their lifetime crosses the jump. To improve the situation, a little trick is
needed. The trace in Figure~\ref{fig:step1} represents a loop, i.e. the jump at the end jumps to
the beginning. Where in the loop the jump occurs is arbitrary, since the loop
can only be left via failing guards anyway. Therefore it does not change the
semantics of the loop to put the jump at another point into the trace and we
can move the \texttt{jump} operation just above the allocation of the objects that
appear in the current \texttt{jump}. This needs some care, because the arguments to
\texttt{jump} are all currently live variables, thus they need to be adapted.

If we do that for our example trace, the trace looks like this:
\begin{figure}
\includegraphics{figures/step2.pdf}
\end{figure}

@@

\section{Supporting Techniques}
\label{sec:support}

XXX virtualizables, store sinking

do we need this?

% section Supporting Techniques (end)
